---
title: "Replica I205 Usage Analysis"
subtitle: "Quantifying Corridor User's Bridge Usage Behavior"
author: "Mike Gaunt"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: TRUE
    toc_float:
      collapsed: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  cache = FALSE, cache.lazy = FALSE, autodep = TRUE, warning = FALSE, 
  message = FALSE, echo = F, dpi = 180,
  fig.width = 5, fig.height = 2, echo = FALSE
  )
```

<!--#general comments===========================================================
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
#
# This is [[insert description here - what it does/solve]]
#
# By: mike gaunt, michael.gaunt@wsp.com
#
# README: [[insert brief readme here]]
#-------- [[insert brief readme here]]
#
# *please use 80 character margins
# *please go to https://pkgs.rstudio.com/flexdashboard/articles/layouts.html
# to explore the different layouts available for the flexdashboard framework
#
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~ -->

<!--#library set-up=============================================================
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
#content in this section should be removed if in production - ok for dev -->
```{r}
library(tidyverse)
library(data.table)
library(here)
library(hms)
library(lubridate)
library(sf)
library(mapview)
library(plotly)
```

<!--#source helpers/utilities===================================================
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
#content in this section should be removed if in production - ok for dev -->

<!--#source data================================================================
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
#content in this section should be removed if in production - ok for dev 
#area to upload data with and to perform initial munging
#please add test data here so that others may use/unit test these scripts -->

## Data Load 
```{r}
portland_pnlact_agg = here("req_portland_205/data"
                           ,"portland_pnlact_agg_20230227.csv") %>%  
  fread() %>%  
  mutate(network_link_ids_unnested = as.character(network_link_ids_unnested ))
```

```{r}
index_link = data.frame(name = c('abernathy_w', "abernathy_e", "tualatin_w", "tualatin_e")
                        ,id = c('13674580744864527942', '16170741915355440261'
                                ,'8441177935916242835', '7566041862080809780'))
```

```{r}
portland_pnlact_agg_pro = portland_pnlact_agg %>%  
  merge(index_link, by.x = "network_link_ids_unnested", by.y = "id")
```

```{r}
trip_table = fread(here::here(here("req_portland_205/data", "trip_table.csv") ))
```

```{r}
rate_df = data.frame(
  start_time = c(0L,1L,2L,3L,4L,5L,6L,7L,8L,9L,10L,
                 11L,12L,13L,14L,15L,16L,17L,18L,19L,20L,21L,22L,
                 23L),
  rate = c(0.55,0.55,0.55,0.55,0.55,1,1.9,1.9,
           1.9,1,0.65,0.65,0.65,1,1,1.65,2.2,2.2,1.65,1,0.65,
           0.65,0.65,0.55)
)
```

```{r, echo=FALSE}
links_t =  here("req_portland_205/data/data_20230222_130855", "replica_queried_network_links.gpkg") %>% 
  read_sf() %>%  
  filter(stableEdgeId %in% c('13674580744864', '16170741915355', '84411779359162', '75660418620808'))

links_a =  here("req_portland_205/data/data_20230222_130430", "replica_queried_network_links.gpkg") %>% 
  read_sf() %>%  
  filter(stableEdgeId %in% c('13674580744864', '16170741915355', '84411779359162', '75660418620808'))

links = bind_rows(
  links_a
  ,links_t
)
```


### Inspect

Bridge links used in query.
```{r}
links %>% mapview(lwd = 16)
```

Sample of query data, each record details an activity_id, person_id, and link attributes.
This table contains duplicate activity_ids if an activity uses more than 1 bridge link.
```{r}
portland_pnlact_agg_pro %>%  
  sample_n(10)
```

Toll rates used for single bridge usage:
```{r}
temp_plot = rate_df %>%  
  ggplot() + 
  geom_col(aes(start_time, rate))

ggplotly(temp_plot)
```

## Analysis Observations

### Activity Based Metrics

Number of unique persons that use at least one bridge:
```{r, echo=FALSE}
portland_pnlact_agg_pro %>% count(person_id) %>%  nrow() 
```

Number of activities that use at least one bridge:
```{r, echo=FALSE}
portland_pnlact_agg_pro %>% count(activity_id) %>%  nrow() 
```

Number of unique bridge usages - calculated 2X ways, should be the same:
```{r, echo=FALSE}
#numbers should be the same
portland_pnlact_agg_pro %>% count() 
#this wont work anymore as count of 2+ indicate it was duplicated across bridges
# portland_pnlact_agg_pro %>% summarise(count = sum(count)) 
```

Number of activities that use [1, 2, ...N] bridge links:
```{r, echo=FALSE}
portland_pnlact_agg_pro %>% 
  count(activity_id, name = "Bridges Used") %>%  
  count(`Bridges Used`, name = "count")  %>%  
  mutate(pct = round(count/sum(count), 2)) 
```

Above table broken out by bridge usage link combinations. 
```{r, echo=F}
summary_bridge_by_activity = portland_pnlact_agg_pro %>%  
  group_by(activity_id) %>%  
  arrange(name) %>%
  mutate(index = row_number()) %>%  
  arrange(desc(index)) %>%  
  ungroup() %>% 
  pivot_wider(id_cols = activity_id, names_from = index,  values_from = name) %>%  
  group_by(`1`, `2`, `3`) %>%
  summarise(count = n()) %>%  
  ungroup() %>%  
  arrange( `3`, `2`)

summary_bridge_by_activity %>%  
  mutate(Bridge_Usage = case_when(
    (is.na(`3`) & is.na(`2`))~"1 Bridges Used"
    ,(is.na(`3`) & !is.na(`2`))~"2 Bridges Used"
    ,T~"3 Bridges Used"
  )) %>%  
  select(Bridge_Usage, count, everything())

```

```{r}
yolo = merge(
  portland_pact_agg %>%  
    group_by(person_id) %>%  
    summarise(activites_bin = n()) 
  ,portland_pact_agg %>%  
    group_by(person_id) %>%  
    summarise(bridge_uses = sum(bridge_usage)))

yolo = yolo %>%  
  group_by(activites_bin, bridge_uses) %>%  
  summarise(persons = n()) %>%  
  ungroup() %>%  
  mutate(percent = round(persons/sum(persons), 2)) %>%  
  mutate(bin_ttl_activites = activites_bin*persons) %>%  
  mutate(bin_ttl_bridge_uses = bridge_uses*persons)

persons_per_activity = yolo %>% 
  group_by(activites_bin) %>%  
  summarise(persons = sum(persons)) %>%  
  ungroup() %>% 
  mutate(percent = 100*round(persons/sum(persons), 2))

persons_per_bridge_uses = yolo %>% 
  group_by(bridge_uses) %>%  
  summarise(persons = sum(persons)) %>%  
  ungroup() %>% 
  mutate(percent = 100*round(persons/sum(persons), 2)
         ,percent_cumm = cumsum(percent))

bridge_uses_per_activity = yolo %>% 
  group_by(activites_bin) %>%  
  summarise(bin_ttl_bridge_uses = sum(bin_ttl_bridge_uses)) %>%  
  ungroup() %>%  
  mutate(percent = 100*round(bin_ttl_bridge_uses/sum(bin_ttl_bridge_uses), 2)
         ,percent_cumm = cumsum(percent))
  
```

Number of persons given activity count.
```{r}
temp_plot = persons_per_activity %>%  
  ggplot() + 
  geom_col(aes(activites_bin, persons ))

persons_per_activity 

ggplotly(temp_plot)
```

Number of persons given bridge usage count.
```{r}
temp_plot = persons_per_bridge_uses %>%  
  ggplot() + 
  geom_col(aes(bridge_uses , persons ))

persons_per_bridge_uses 

ggplotly(temp_plot)
```

Bridges used given activity count.
```{r}
temp_plot = bridge_uses_per_activity %>%  
  ggplot() + 
  geom_col(aes(activites_bin , bin_ttl_bridge_uses ))

bridge_uses_per_activity 

ggplotly(temp_plot)
```

### Person Based Metrics

```{r}
#calculate bridge usage per person and activity
portland_pact_agg = portland_pnlact_agg_pro %>% 
  group_by(activity_id, person_id) %>%  
  summarise(bridge_usage = n()) %>%  
  ungroup() 

portland_pact_agg_pro = portland_pact_agg %>%  
  group_by(person_id) %>%  
  mutate(person_activities = n()) %>%  
  ungroup() %>%  
  mutate(person_activities_flag = case_when(
    person_activities == 1~ "1"
    ,person_activities != 1~ "2+"
    ,T~"PROBLEM"
  ))

```

```{r}
#process trip_table
trip_table_por = trip_table %>%  
  select(activity_id, person_id, start_time, end_time, tour_type) %>%  
  mutate(start_time = hms::as_hms(start_time)
         ,end_time = hms::as_hms(end_time)
         ,middle_time = hms::as_hms(start_time+(end_time-start_time)/2)) %>%  
  mutate(start_time_h = hour(start_time)
         ,end_time_h = hour(end_time)
         ,middle_time_h = hour(middle_time))
```

Distribution of activity start times.
```{r}
temp_plot = trip_table_por %>%  
  mutate(count = 1) %>%  
  group_by(tour_type, start_time_h) %>%  
  summarise(count = sum(count)) %>% clipr::write_clip()
  ggplot() + 
  geom_col(aes(start_time_h, count, fill = tour_type), position = "dodge")

ggplotly(temp_plot)
```

Distribution of start time differences between subsequent trips.
```{r}
temp_data = trip_table_por %>%  
  arrange(person_id, start_time) %>%  
  group_by(person_id) %>%  
  filter(n()!=1) %>% 
  mutate(start_diff = as.numeric(start_time-lag(start_time))/3600) 

temp_plot = temp_data %>%  
  mutate(start_time_difference = start_diff %>%  
           gauntlet::floor_divide(1)
         ,count = 1) %>% 
  group_by(tour_type, start_time_difference) %>%  
  summarise(count = sum(count)) %>% clipr::write_clip()
  ggplot() + 
  geom_col(aes(start_time_difference, count, fill = tour_type))

ggplotly(temp_plot)
```

#### Cost Projections
```{r}
#merge trip table to portland_pact_agg and rate
merged_trip_count_cost = trip_table_por %>%  
  select(!c(start_time, end_time, middle_time)) %>% 
  merge(portland_pact_agg
        ,by = c("activity_id", "person_id")) %>%  
  pivot_longer(cols = ends_with("_h")
               ,names_to = "analysis"
               ,values_to = "hour") %>% 
  merge(rate_df
        ,by.x = "hour", by.y = "start_time") %>%  
  mutate(charged_amount = bridge_usage*rate)

#aggregate to person 
merged_trip_count_cost_agg = merged_trip_count_cost %>%  
  group_by(analysis, person_id) %>%  
  summarise(bridge_usage_ttl = sum(bridge_usage)
            ,charged_amount_ttl = sum(charged_amount))
```

Distribution of activity toll costs.
```{r}
merged_trip_count_cost %>%  
  mutate(count = 1) %>%  
  ggplot() + 
  geom_histogram(aes(charged_amount, fill = analysis), position = "dodge" )
```

Table of percent of persons paying less than or equal to $4.4.
```{r}
merged_trip_count_cost_agg %>%  
  group_by(analysis) %>%  
  summarise(percent_less_than_4.4 = mean(charged_amount_ttl<=4.4)) 
```

Distribution of total daily toll cost per person.
```{r}
temp_plot = merged_trip_count_cost_agg %>%  
  filter(analysis == "end_time_h") %>%  
  arrange(desc(charged_amount_ttl)) %>% 
  mutate(count = 1) %>%  
  ggplot() + 
  geom_histogram(aes(charged_amount_ttl, fill = analysis), position = "dodge" )

ggplotly(temp_plot)
```

Distribution of average bridge toll cost per person.
```{r}
temp_plot = merged_trip_count_cost_agg %>%  
  filter(analysis == "end_time_h") %>%  
  arrange(desc(charged_amount_ttl)) %>% 
  mutate(count = 1) %>%  
  ggplot() + 
  geom_histogram(aes(charged_amount_ttl/bridge_usage_ttl, fill = analysis),position = "dodge" )

ggplotly(temp_plot)
```

#### Cost Projections Split
```{r}
#merge trip table to portland_pact_agg and rate
merged_trip_count_cost = trip_table_por %>%  
  select(!c(start_time, end_time, middle_time)) %>% 
  merge(portland_pact_agg_pro
        ,by = c("activity_id", "person_id")) %>% 
  pivot_longer(cols = ends_with("_h")
               ,names_to = "analysis"
               ,values_to = "hour") %>% 
  merge(rate_df
        ,by.x = "hour", by.y = "start_time") %>%  
  mutate(charged_amount = bridge_usage*rate)

#aggregate to person 
merged_trip_count_cost_agg_act = merged_trip_count_cost %>%  
  filter(analysis == "start_time_h") %>% 
  group_by(person_id, person_activities) %>%  
  summarise(bridge_usage_ttl = sum(bridge_usage)
            ,charged_amount_ttl = sum(charged_amount))
```

Distribution of activity toll costs.
```{r}
#cost per trip 
#THIS CHECKS OUT!!!!!!!!!!!!!!!!
temp_1 = merged_trip_count_cost %>%  
  mutate(charged_amount_ttl_bin = gauntlet::floor_divide(charged_amount, .5)
         ,count = 1) %>%  
  group_by(charged_amount_ttl_bin) %>%
  summarise(count = sum(count)) %>%  
  ungroup() 

temp_1 %>%  
  merge(
    data.frame(
      charged_amount_ttl_bin = seq(0, max(temp_1$charged_amount_ttl_bin), .5)
    ), all.y = T) %>%  
  mutate(count = replace_na(count, 0)) 

#PLOT: ACTIVITIES PER COST BIN 
ggplot(temp_1
       ,aes(charged_amount_ttl_bin, count)) +  
  geom_col() 

#cost per trip 
#THIS CHECKS OUT!!!!!!!!!!!!!!!!
temp_2 = merged_trip_count_cost %>%  
  filter(analysis == "start_time_h") %>% 
  mutate(charged_amount_ttl_bin = gauntlet::floor_divide(charged_amount, .5)
         ,count = 1) %>%  
  group_by(tour_type, charged_amount_ttl_bin) %>%
  summarise(count = sum(count)) %>%  
  ungroup() 

temp_2 %>%  
  merge(
    data.frame(
      charged_amount_ttl_bin = seq(0, max(temp_1$charged_amount_ttl_bin), .5)
    ), all.y = T) %>%  
  complete(tour_type, charged_amount_ttl_bin) %>% 
  mutate(count = replace_na(count, 0)) 

#PLOT: ACTIVITIES PER COST BIN BY TOUR TYPE
ggplot(temp_2
       ,aes(charged_amount_ttl_bin, count, fill = tour_type)) +  
  geom_col() 

```

Table of percent of persons paying less than or equal to $4.4.
```{r}
merged_trip_count_cost_agg_act %>%  
  group_by( person_activities_flag ) %>%  
  summarise(percent_less_than_4.4 = mean(charged_amount_ttl<=4.4)) %>% 
  clipr::write_clip()
```

Distribution of total daily toll cost per person.
```{r}
temp_1_p = merged_trip_count_cost_agg_act %>%  
  mutate(charged_amount_ttl_bin = gauntlet::floor_divide(charged_amount_ttl, .1)) %>% 
  group_by(person_activities, charged_amount_ttl_bin) %>%  
  summarise(count = n())

temp_1_prp = temp_1_p %>% 
  merge(
    .
    ,data.frame(
      charged_amount_ttl_bin = seq(0, max(temp_1_p$charged_amount_ttl_bin), .1)
    ), all.y = T) %>%  
  complete(person_activities, charged_amount_ttl_bin) %>% 
  mutate(count = replace_na(count, 0)) %>%  
  clipr::write_clip()



temp_1_p = merged_trip_count_cost_agg_act %>%  
  mutate(charged_amount_ttl_bin = gauntlet::floor_divide(charged_amount_ttl, .1)) %>% 
  group_by(charged_amount_ttl_bin) %>%  
  summarise(count = n())

temp_1_prp = temp_1_p %>% 
  merge(
    .
    ,data.frame(
      charged_amount_ttl_bin = seq(0, max(temp_1_p$charged_amount_ttl_bin), .1)
    ), all.y = T) %>%  
  # complete(person_activities_flag , charged_amount_ttl_bin) %>% 
  mutate(count = replace_na(count, 0)) %>%  
  clipr::write_clip()





```

Distribution of average bridge toll cost per person.
```{r}
temp_plot = merged_trip_count_cost_agg %>%  
  filter(analysis == "end_time_h") %>%  
  arrange(desc(charged_amount_ttl)) %>% 
  mutate(count = 1) %>%  
  ggplot() + 
  geom_histogram(aes(charged_amount_ttl/bridge_usage_ttl, fill = analysis),position = "dodge" )

ggplotly(temp_plot)
```



```{r}
portland_pact_agg_pro_3 %>%  
  mutate
```



```{r}
merged_trip_count_cost_agg %>%  
  group_by(analysis,
           cost_bin = cut(charged_amount_ttl, c(0:10, 100), right = F)) %>% 
  summarise(count = n()) %>%  
  group_by(analysis) %>%  
  mutate(pct = count/sum(count)) %>%  
  ungroup() %>%  
  ggplot() + 
  geom_col(aes(cost_bin, count, fill = analysis), position = "dodge" )

merged_trip_count_cost_agg %>%  
  group_by(analysis, thresh_4 = charged_amount_ttl > 4) %>%  
  summarise(count = n())


```


```{r}
merged_trip_count_cost = trip_table_por %>%  
  merge(portland_pact_agg
        ,by = c("activity_id", "person_id")) %>%  
  pivot_longer(cols = ends_with("time")
               ,names_to = "analysis"
               ,values_to = "hour") %>%  
  merge(rate_df
        ,by.x = "hour", by.y = "start_time") %>%  
  mutate(charged_amount = bridge_usage*rate)

merged_trip_count_cost_agg = merged_trip_count_cost %>%  
  group_by(analysis, person_id) %>%  
  summarise(bridge_usage_ttl = sum(bridge_usage)
            ,charged_amount_ttl = sum(charged_amount))


merged_trip_count_cost_agg %>%  
  mutate(count = 1) %>%  
  filter(charged_amount_ttl < 10) %>% 
  ggplot() + 
  geom_histogram(aes(charged_amount_ttl, fill = analysis), position = "dodge" )


merged_trip_count_cost_agg %>%  
  group_by(analysis,
           cost_bin = cut(charged_amount_ttl, c(0:10, 100), right = F)) %>% 
  summarise(count = n()) %>%  
  group_by(analysis) %>%  
  mutate(pct = count/sum(count)) %>%  
  ungroup() %>%  
  ggplot() + 
  geom_col(aes(cost_bin, count, fill = analysis), position = "dodge" )

merged_trip_count_cost_agg %>%  
  group_by(analysis, thresh_4 = charged_amount_ttl > 4) %>%  
  summarise(count = n())


```



<!--end-->






